### 2013-05-02 doc/dissertation/document/chapters/mc_background/mc_background.tex
some clean up in the mc background section

In this chapter, we will present the fundamentals of the Monte Carlo
method for discrete linear systems. Both the forward and adjoint
methods will be presented and analyzed including a variance analysis
of several Monte Carlo estimators. The Sequential Monte Carlo and
Monte Carlo Synthetic Acceleration methods will then be presented as a
means of leveraging the basic Monte Carlo methods in an iterative
refinement scheme that accelerates their convergence. Using these
iterative schemes, a set of modest variance reduction techniques will
be introduced and their effects on the Monte Carlo methods explored.

@@

\subsection{Direct Method: Evolution of a Solution}
\label{subsec:direct_evolution}
As a means of visually demonstrating the direct Monte Carlo method,
consider a 2-dimensional thermal diffusion problem with sources on the
left and right hand sides of the domain and a uniform source
throughout the domain of 1/5 the strength of the boundary sources as
shown in Figure~\ref{fig:heat_setup}.
\begin{figure}[p!]
  \begin{center}
    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for 2D heat equation.}
    \textit{Dirichlet conditions are set for the temperature on all 4
      boundaries of the Cartesian grid. Background source of 1/5 the
      value of the boundary sources present. $50 \times 50$ grid.}}
  \label{fig:heat_setup}
\end{figure}
For this problem, the number of histories used to compute the solution
at each grid point (state) in the domain was increased from 1 to 1000
in order to demonstrate the effects on the solution and the
statistical nature of the method. Figure~\ref{fig:direct_evolution}
gives these results.
\begin{figure}[p!]
  \begin{center}
    \includegraphics[width=5in]{chapters/mc_background/direct_evolution.png}
  \end{center}
  \caption{\textbf{Direct Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 1 history
      per state. Top right: 10 histories per state. Bottom left: 100
      histories per state. Bottom right: 1000 histories per state.}}
  \label{fig:direct_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tallies are
made. Starting with single history at each state in the
domain, the high variance prevents a precise solution from being
obtained although we begin to see the solution take shape as
expected. At 1000 histories per state, enough tallies have been made
to generate a reasonable estimate for the structure of the
solution. It is interesting to note here that as the statistical
uncertainty is reduced at each grid point in the domain, the solution
is resolved in a certain sense, analagous to the convergence of a
traditional iterative method.

%%---------------------------------------------------------------------------%%
\section{Adjoint Monte Carlo Method}
\label{sec:adjoint_mc}
Often a more useful form, an alternative to forward Monte Carlo matrix
inversion is the adjoint method. We begin by defining the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively vectors and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
mirror Eq~(\ref{eq:richardson_split}): 
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. By defining the following
inner product equivalence \citep{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

Expanding the transition probabilities in the first sum again yields a
Neumman series in the same form as that explored for the forward
estimator, this time with the terms in reverse order and the
introduction of the Kronecker delta:
\begin{equation}
  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
  b^2_{i_0}\delta_{i_k,j} - x_j^2\:.
  \label{eq:collision_variance_2}
\end{equation}
If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
then have:
\begin{equation}
  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} - x_j^2\:,
\label{eq:collision_variance_3}
\end{equation}
where now $\ve{G}$ in this case is the transpose of that used in
Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
that the variance contribution from each random walk will only be in
its current state and for many random walks, many starting weights of
$b_{i_0}$ will contribute to the variance in the same way as they
contribute to the solution tally. As the transpose is formed and the
Neumann series of $\ve{G}$ is in reverse order in
Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
collision estimator is bound by the same restrictions on
$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
finite.

@@

\subsubsection{Expected Value Estimator Variance}
\label{subsubsec:expected_value_estimator_variance}
Following the collision estimator variance, the expected value
estimator variance is given as:
\begin{multline}
  \sigma_j^2 = \sum_{\nu} P_{\nu} b_j^2 + 2 \sum_{\nu} P_{\nu} b_j
  \sum_{m=0}^k W_{m} h_{j,i_m} +\\ \sum_{\nu} P_{\nu} \sum_{m=0}^k
  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{\substack{ l=0
      \\ m<l}}^k W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
  \label{eq:expected_value_variance_1}
\end{multline}
As the delta functions are no longer present, the final summation in
Eq~(\ref{eq:expected_value_variance_1}) contains contributions for all
valid state combinations in the same and therefore the variance of
each state in the solution tally is coupled to a group of other states
as defined by the sparsity pattern of the iteration. This coupling of
variances is expected due to the deterministic averaging used to
generate the expected value estimator.

\subsection{Adjoint Method: Evolution of a Solution}
\label{subsec:adjoint_evolution}
As a means of visually demonstrating the adjoint Monte Carlo method,
again consider the 2-dimensional thermal diffusion problem with
sources on the left and right hand sides of the domain and a smaller
uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
adjoint method with the collision estimator, the number of histories
sampled from the source was increased from 10 to 10,000,000 in order
to show its effects on the solution and the statistical nature of the
method. Figure~\ref{fig:adjoint_evolution} gives these results.
\begin{figure}[p!]
  \begin{center}
    \includegraphics[width=5in]{chapters/mc_background/adjoint_evolution.png}
  \end{center}
  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 10
      histories per state. Top right: 1,000 histories per
      state. Bottom left: 100,000 histories per state. Bottom right:
      10,000,000 histories per state.}}
  \label{fig:adjoint_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tally
contributions are made. At 10,000,000 histories per state, enough
information has been tallied to generate a reasonable estimate for the
structure of the solution. The visual difference between
Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
precisely that determined by their mathematics. As the adjoint
solution evolves with the addition of histories, more histories
emanate from the boundary and smaller uniform source with more
penetrating from the boundary into the domain and making contributions
to the tallies in those states.

%%---------------------------------------------------------------------------%%
\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \citep{halton_sequential_1962} and further refined
his work some years later \citep{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $exp(-N)$ \citep{evans_residual_2003}. In
much the same way as projection methods, Halton's method, sequential
Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
directly reducing the elements residual vector. He proposed the
following iterative scheme as a solution to
Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}^k$ is computed by the
adjoint Monte Carlo method at each iteration. The merits of Halton's
approach are immediately visible in that we have now broken the
binding of the convergence rate to the Central Limit Theorem. Here,
the Monte Carlo solver is used to produce a correction from the
residual, analogous to using the residual to extract a correction from
the search subspace in a projection method. By doing this, the Monte
Carlo error is bound in the correction used to update the solution and
therefore does not explicitly manifest itself in the overall
convergence of the solution. The downside of such a method is that if
the solution guess is poor, then many iterations are required in order
to reach exponential converge as the Monte Carlo error (and therefore
the Central Limit Theorem) does dominate in this situation.

@@

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method with the collision estimator as
compared to the direct method. To assess both the CPU time and number
of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of the number of mesh
elements, fixing the spectral radius of the system at a constant value
for each variation. Both the five-point and nine-point stencils were
used with both the direct and adjoint solvers. For each case, $N
\times N$ total random walk permutations were computed per MCSA
iteration where $N \times N$ is the number of discrete grid points in
the system. Solver parameters were set to a weight cutoff of
\sn{1}{-4} for the stochastic linear solver and a convergence
tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance as a function of the problem size. All
computations presented in this section and
\S\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel
Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the adjoint and direct
      solvers are used with the five-point and nine-point stencils.}}
  \label{fig:poisson_iterations}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
gives the results of these computations. First, it is worthy to note
on the semilog plot that we are indeed achieving the expected
exponential convergence from MCSA with both Monte Carlo
solvers. Second, we note that using the adjoint method with the same
number of stochastic histores per MCSA iteration gives a faster rate
of converge for the same reasons as above. We also note here that
fewer iterations are required for convergence when the 9-point stencil
is used to discretize the Laplacian operator (although at no gain in
speed as given by the results in
Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
smaller discretization error directly corresponds to a more well
defined residual source generated by the Richardson extrapolation for
the Monte Carlo calculation. In addition, the better defined source is
transported through a domain described more accurately by the 9-point
stencil, thus yielding a more accuracte correction vector from the
Monte Carlo calculation.

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the adjoint and direct solvers are used with the five
      point and nine point stencils. A higher rate of convergence is
      observed for MCSA using the adjoint Monte Carlo solver as
      compared to the direct method when both solvers compute the same
      number of random walks per iteration.}}
  \label{fig:poisson_convergence}
\end{figure}

@@

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/seq_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver. The number of random walks was
      twice the number of discrete states in the system in order to
      ensure convergence in the Sequential Monte Carlo method.}}
  \label{fig:seq_cpu_time}
\end{figure}

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/seq_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver.}}
  \label{fig:seq_iterations}
\end{figure}

@@

\subsection{Weight Capping}
\label{subsec:weight_capping}

@@

>>>>

In this chapter, we will present the fundamentals of the Monte Carlo
method for discrete linear systems. Both the forward and adjoint
methods will be presented and analyzed including a brief variance
analysis of several Monte Carlo estimators. The Sequential Monte Carlo
and Monte Carlo Synthetic Acceleration methods will then be presented
as a means of leveraging the basic Monte Carlo methods in an iterative
refinement scheme that accelerates their convergence. Using these
iterative schemes, a set of modest variance reduction techniques will
be introduced and their effects on the Monte Carlo methods explored.

@@

\subsection{Direct Method: Evolution of a Solution}
\label{subsec:direct_evolution}
As a means of visually demonstrating the direct Monte Carlo method,
consider a 2-dimensional thermal diffusion problem with sources on the
left and right hand sides of the domain and a uniform source
throughout the domain of 1/5 the strength of the boundary sources as
shown in Figure~\ref{fig:heat_setup}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for 2D heat equation.}
    \textit{Dirichlet conditions are set for the temperature on all 4
      boundaries of the Cartesian grid. Background source of 1/5 the
      value of the boundary sources present. $50 \times 50$ grid.}}
  \label{fig:heat_setup}
\end{figure}
For this problem, the number of histories used to compute the solution
at each grid point (state) in the domain was increased from 1 to 1000
in order to demonstrate the effects on the solution and the
statistical nature of the method. Figure~\ref{fig:direct_evolution}
gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/mc_background/direct_evolution.png}
  \end{center}
  \caption{\textbf{Direct Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 1 history
      per state. Top right: 10 histories per state. Bottom left: 100
      histories per state. Bottom right: 1000 histories per state.}}
  \label{fig:direct_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tallies are
made. Starting with single history at each state in the
domain, the high variance prevents a precise solution from being
obtained although we begin to see the solution take shape as
expected. At 1000 histories per state, enough tallies have been made
to generate a reasonable estimate for the structure of the
solution. It is interesting to note here that as the statistical
uncertainty is reduced at each grid point in the domain, the solution
is resolved in a certain sense, analagous to the convergence of a
traditional iterative method.
\clearpage

%%---------------------------------------------------------------------------%%
\section{Adjoint Monte Carlo Method}
\label{sec:adjoint_mc}
Often a more useful form, an alternative to forward Monte Carlo matrix
inversion is the adjoint method. We begin by defining the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
mirror Eq~(\ref{eq:richardson_split}):
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. By defining the following
inner product equivalence \citep{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

Expanding the transition probabilities in the first sum again yields a
Neumman series in the same form as that explored for the forward
estimator, this time with the terms in reverse order and the
introduction of the Kronecker delta:
\begin{multline}
  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{\substack{
      l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
  \label{eq:collision_variance_2}
\end{multline}
If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
then have:
\begin{multline}
  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu}
  P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j}
  \delta_{i_l,j} - x_j^2\:,
\label{eq:collision_variance_3}
\end{multline}
where now $\ve{G}$ in this case is the transpose of that used in
Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
that the variance contribution from each random walk will only be in
its current state and for many random walks, many starting weights of
$b_{i_0}$ will contribute to the variance in the same way as they
contribute to the solution tally. As the transpose is formed and the
Neumann series of $\ve{G}$ is in reverse order in
Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
collision estimator is bound by the same restrictions on
$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
finite.

@@

\subsubsection{Expected Value Estimator Variance}
\label{subsubsec:expected_value_estimator_variance}
Following the collision estimator variance, the expected value
estimator variance is given as:
\begin{multline}
  \sigma_j^2 = \sum_{\nu} P_{\nu} b_j^2 + 2 \sum_{\nu} P_{\nu} b_j
  \sum_{m=0}^k W_{m} h_{j,i_m} +\\ \sum_{\nu} P_{\nu} \sum_{m=0}^k
  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{\substack{ l=0
      \\ m<l}}^k W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
  \label{eq:expected_value_variance_1}
\end{multline}
As the delta functions are no longer present, the final summation in
Eq~(\ref{eq:expected_value_variance_1}) contains contributions for all
valid state combinations in the same row of the iteration matrix and
therefore the variance of each state in the solution tally is coupled
to a group of other states as defined by the sparsity pattern of the
iteration matrix. This coupling of variances is expected due to the
deterministic averaging used to generate the expected value estimator.

\subsection{Adjoint Method: Evolution of a Solution}
\label{subsec:adjoint_evolution}
As a means of visually demonstrating the adjoint Monte Carlo method,
again consider the 2-dimensional thermal diffusion problem with
sources on the left and right hand sides of the domain and a smaller
uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
adjoint method with the collision estimator, the number of histories
sampled from the source was increased from 10 to 10,000,000 in order
to show its effects on the solution and the statistical nature of the
method. Figure~\ref{fig:adjoint_evolution} gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/mc_background/adjoint_evolution.png}
  \end{center}
  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 10
      histories per state. Top right: 1,000 histories per
      state. Bottom left: 100,000 histories per state. Bottom right:
      10,000,000 histories per state.}}
  \label{fig:adjoint_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tally
contributions are made. At 10,000,000 histories per state, enough
information has been tallied to generate a reasonable estimate for the
structure of the solution. The visual difference between
Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
precisely that determined by their mathematics. As the adjoint
solution evolves with the addition of histories, more histories
emanate from the boundary and smaller uniform source with more
penetrating from the boundary into the domain and making contributions
to the tallies in those states.
\clearpage

%%---------------------------------------------------------------------------%%
\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \citep{halton_sequential_1962} and further refined
his work some years later \citep{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In
much the same way as projection methods, Halton's method, sequential
Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
directly reducing the elements residual vector. He proposed the
following iterative scheme as a solution to
Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}^k$ is computed by the
adjoint Monte Carlo method at each iteration. The merits of Halton's
approach are immediately visible in that we have now broken the
binding of the convergence rate to the Central Limit Theorem. Here,
the Monte Carlo solver is used to produce a correction from the
residual, analogous to using the residual to extract a correction from
the search subspace in a projection method. By doing this, the Monte
Carlo error is bound in the correction used to update the solution and
therefore does not explicitly manifest itself in the overall
convergence of the solution. The downside of such a method is that if
the solution guess is poor, then many iterations are required in order
to reach exponential converge as the Monte Carlo error (and therefore
the Central Limit Theorem) does dominate in this situation.

@@

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method with the collision estimator as
compared to the direct method. To assess both the CPU time and number
of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of the number of mesh
elements, fixing the spectral radius of the system at a constant value
for each variation. Both the five-point and nine-point stencils were
used with both the direct and adjoint solvers. For each case, $N
\times N$ total random walk permutations were computed per MCSA
iteration where $N \times N$ is the number of discrete grid points in
the system. Solver parameters were set to a weight cutoff of
\sn{1}{-4} for the stochastic linear solver and a convergence
tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance as a function of the problem size. All
computations presented in this section and
\S\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel
Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the adjoint and direct
      solvers are used with the five-point and nine-point stencils.}}
  \label{fig:poisson_iterations}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
gives the results of these computations. First, it is worthy to note
on the semilog plot that we are indeed achieving the expected
exponential convergence from MCSA with both Monte Carlo
solvers. Second, we note that using the adjoint method with the same
number of stochastic histores per MCSA iteration gives a faster rate
of converge for the same reasons as above. We also note here that
fewer iterations are required for convergence when the 9-point stencil
is used to discretize the Laplacian operator (although at no gain in
speed as given by the results in
Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
smaller discretization error directly corresponds to a more well
defined residual source generated by the Richardson extrapolation for
the Monte Carlo calculation. In addition, the better defined source is
transported through a domain described more accurately by the 9-point
stencil, thus yielding a more accuracte correction vector from the
Monte Carlo calculation.
\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the adjoint and direct solvers are used with the five
      point and nine point stencils. A higher rate of convergence is
      observed for MCSA using the adjoint Monte Carlo solver as
      compared to the direct method when both solvers compute the same
      number of random walks per iteration.}}
  \label{fig:poisson_convergence}
\end{figure}

\clearpage

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/seq_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver. The number of random walks was
      twice the number of discrete states in the system in order to
      ensure convergence in the Sequential Monte Carlo method.}}
  \label{fig:seq_cpu_time}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/seq_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver.}}
  \label{fig:seq_iterations}
\end{figure}

@@

@@

\clearpage
